Baseline centering problem

### 2018-12-13 paper/tangotd.tex

We propose to learn two functions $\green(s)$ and $f(s)$ from states into
a state representation $E$, such that for any test function $\psi\from
S\to \R$, for any $s_1\in S$,
\begin{equation}
\label{eq:greenrep}
\E_{s_2\sim \rho}\,\green(s_1)\cdot
f(s_2)\psi(s_2)\approx \transp{\1_{s_1}}\,\Delta^{-1} \psi
\end{equation}

If the state space is finite and if we use a one-hot representation, a
solution is indeed $E=\R^{\#S}$ with
\begin{equation}
f(s_2)=\1_{s_2},\qquad
(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)
\end{equation}
but with infinite or continuous state spaces, we have to find a more
compact representation.

The reason for this form is the following: $E$ is supposed to represent
distributions/functions over states. $f(s_2)$ is supposed to be
mapped to $\1_{s_2}$. $\green(s_1)$ is supposed to be mapped to the
density wrt $\rho$ of the fundamental matrix at $s_1$ (the occupation
distribution from $s_1$: $\transp{\1_{s_1}}\sum_t P^t$, where as usual I
omit constants\footnote{To deal with constants, we probably have to be
able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$, or just use
relative value functions and remove $\psi(s_0)$ for some $s_0$.}). In
particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
the vector with components $\rho \psi(s_2)$. Then the scalar product in
$E$ represents integration/summation with respect to $\rho$.

Of course we cannot directly learn $\green$ from \eqref{eq:greenrep},
since $\Delta^{-1}$ in the right-hand-side is exactly what we want to
learn. But we can use the Bellman equations \eqref{eq:greenbell}; after
applying \eqref{eq:greenbell} to an arbitrary state $s_1$ and test
function $\psi$, and
substituting the model $\green$ for $\Delta^{-1}$ via
$\eqref{eq:greenrep}$, the left Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\Delta^{-1}\psi=\transp{\1_{s_1}}\psi+\transp{\1_{s_1}}P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}\,\green(s_1)\cdot
f(s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\E_{s_2\sim
\rho}\,\green(s')\cdot
f(s_2)\psi(s_2)
\end{gather}
because $\transp{\1_{s_1}}P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
Meanwhile,
the right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\Delta^{-1}\psi=\transp{\1_{s_1}}\psi+\transp{\1_{s_1}}\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}\,\green(s_1)\cdot
f(s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
\,\green(s_1)\cdot
f(s_2)\psi(s'_2)
\end{gather}
(indeed, 
$P\psi$ is
represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s_2)$).

@@

The value function is obtained with $\psi=R$, namely,
\begin{equation}
V(s_1)=\transp{\1_{s_1}}\Delta^{-1}R\approx
\E_{s_2\sim \rho}\,\green(s_1)\cdot
f(s_2)R(s_2)
\end{equation}
Of course it is not convenient to have to sample from $s_2$. Instead one
might want to just learn the representation $f(R)$, defined by abuse of
notation as
\begin{equation}
f(R)\deq \E_{s_2\sim \rho} \,f(s_2)R(s_2)
\end{equation}
(which can be computed online)
and then
\begin{equation}
V(s_1)\approx \green(s_1)\cdot f(R)
\end{equation}

>>>>

We propose to learn two functions $\green(s)$ and $f(s)$ from states into
a state representation $E$, such that for any test function $\psi\from
S\to \R$, for any $s_1\in S$,
\begin{equation}
\label{eq:greenrep}
\green(s_1)\cdot \E_{s_2\sim \rho}
f(s_2)\psi(s_2)\approx \transp{\1_{s_1}}\,\Delta^{-1} \psi
\end{equation}

If the state space is finite, a 
solution is indeed $E=\R^{\#S}$ with a one-hot representation for $f$,
and the true matrix $\Delta^{-1}$ for $\green$, namely,
\begin{equation}
f(s_2)=\1_{s_2}/\rho(s_2),\qquad
(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}
\end{equation}
or equivalently, $f(s_2)=\1_{s_2}$ and
$(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)$. \footnote{The solution to
such representation problems is never unique because $E$ is an arbitrary
space and one may always make a change of basis in $E$, e.g., dividing
the $s_2$-coordinate by $\rho(s_2)$.}
But with infinite or continuous state spaces, we have to find a more
compact representation.

The reason for this form is the following: $E$ is supposed to represent
distributions or functions over states. $f$ is supposed to map $s_2$ to
something like
$\1_{s_2}$ (up to a factor $1/\rho$ to compensate for sampling
probabilities). 
In
particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
the vector with components $\psi(s_2)$, namely, the function $\psi$. $\green$ is supposed to map each
$s_1$ to the
the occupation
distribution of the random walk starting from $s_1$, namely,
$\transp{\1_{s_1}}\sum_t P^t$, which is $\transp{\1_{s_1}}\,\Delta^{-1}$
(as usual I
omit constants\footnote{To deal with constants, we probably have to be
able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$, or just use
relative value functions and remove $\psi(s_0)$ for some $s_0$.}). Then the scalar product in
$E$ represents integration/summation over the state space.

Of course we cannot directly learn $\green$ from \eqref{eq:greenrep},
since $\Delta^{-1}$ in the right-hand-side is exactly what we want to
learn. But we can use the Bellman equations \eqref{eq:greenbell}; after
applying \eqref{eq:greenbell} to an arbitrary state $s_1$ and test
function $\psi$, and
substituting the model $\green$ for $\Delta^{-1}$ via
$\eqref{eq:greenrep}$, the left Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\green(s_1)\cdot\E_{s_2\sim \rho}
f(s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\green(s')\cdot
\E_{s_2\sim \rho}\,
f(s_2)\psi(s_2)
\end{gather}
because $\transp{\1_{s_1}}\,P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
Meanwhile,
the right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\green(s_1)\cdot\E_{s_2\sim \rho}
f(s_2)\psi(s_2)
\approx \psi(s_1)+ \green(s_1)\cdot\E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
f(s_2)\psi(s'_2)
\end{gather}
(indeed, 
$P\psi$ is
represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s'_2)$).

@@

Thus we may learn $\green$ via TD-like algorithms. We may even have
double sampling that works, thanks to featurization (using test functions
$\psi$), at least for the right equation. TODO should we featurize on the
left as well?

The value function is obtained with $\psi=R$, namely,
\begin{equation}
V(s_1)=\transp{\1_{s_1}}\Delta^{-1}R\approx
\green(s_1)\cdot\E_{s_2\sim \rho}
f(s_2)R(s_2)
\end{equation}
Of course it is not convenient to have to sample from $s_2$. Instead one
might want to just learn the representation $f(R)$, defined by abuse of
notation as
\begin{equation}
f(R)\deq \E_{s_2\sim \rho} \,f(s_2)R(s_2)
\end{equation}
(which can be computed online)
and then
\begin{equation}
V(s_1)\approx \green(s_1)\cdot f(R)
\end{equation}
